---
title: "Approach II Part 2"
author: "Astha Bista"
date: "2/15/2021"
output: md_document
bibliography: [packages.bib,ref.bib]
biblio-style: "apalike"
link-citations: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r , warning = FALSE, include=FALSE, echo=FALSE}
list.of.packages <- c('TSclust')
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
library('TSclust')

```

Here is a plot comparing the decadal and seasonal oscillations in the variables of the subgroup of W70 well.
```{r echo=FALSE}
variableList <- list(c("W60","G42","Kearney (KY) precipitation","Kearney (KY) temperature","Pumping"),
                     c("W63","G42","Kearney (KY) precipitation","Kearney (KY) temperature","Pumping"),
                     c("W67","G45","Kearney (KY) precipitation","Kearney (KY) temperature","Pumping"),
                     c("W70","G40","Canaday steam plant (CDY) precipitation","Canaday steam plant (CDY) temperature","Pumping"), c("W73","G40","Canaday steam plant (CDY) precipitation","Canaday steam plant (CDY) temperature","Pumping"),
                     c("W74","G42","Gothenberg (GO) precipitation","Gothenberg (GO) temperature","Pumping"),
                     c("W78","G49","Grand Island (GI) precipitation","Grand Island (GI) temperature","Pumping"),
                     c("W80","G42","Gothenberg (GO) precipitation","Gothenberg (GO) temperature","Pumping"),
                     c("W81","G49","Grand Island (GI) precipitation","Grand Island (GI) temperature","Pumping"),
                     c("W115","G53","Grand Island (GI) precipitation","Grand Island (GI) temperature","Pumping"),
                     c("W116","G53","Grand Island (GI) precipitation","Grand Island (GI) temperature","Pumping"),
                     c("W118","G65","David City (DC) precipitation","David City (DC) temperature","Pumping"))

# List of seasonal and decadal variations in MSSA operations for each respective subgroup
var_explained <- list(c(36.08,33.92,35.65,37.84,38.53,36.01,38.47,37.67,38.66,37.72,37.93,38.06),
                      c(18.19,27.76,23.62,24.8,24.34,19.57,23.8,17.64,23.69,23.35,21.71,15.8))
osc <- c("Seasonal","Decadal")

list <- c("W","G","P","T","Pu")
```

```{r fig.height=5,fig.width=15}

par(mfrow = c(1, 2))

## Loop around seasonal and decadal oscillations
for (j in 1:2){
    # Assign filepath for each subgroup
    assign(paste0("filepath",j),Sys.glob(path=paste0("Approach_II/MSSA_Work/W70_RC",
                                                     sprintf("%02d",as.numeric(j)),".csv")))
    # Read RC files for decadal and seasonal in each subgroup
    assign(paste0("F",j),read.csv(file=get(paste0("filepath",j)),header = TRUE))
    
    ## Loop around each variable 
    for (k in 1:length(list)){
        # Create time series object for each variable
        assign(paste0(list[k],".ts"),ts(get(paste0("F",j))[,k+1],start=c(2000,1),end=c(2018,12),
                                        frequency = 12))
    }
    
    ## Plot time series of all variables for decadal and seasonal separately
    ts.plot(W.ts,G.ts,P.ts,T.ts,Pu.ts,
            gpars = list(xlab="Year",ylab="Standardized units",xaxt="n",ylim=c(-2,4),
                         col=c("green","blue","purple","orange","dark green"),lty=c(1,1,2,2,2,3),lwd=c(2,2,2,2,2,3)))
    
    year.text=c("2000","2001","2002","2003","2004","2005","2006","2007",
                "2008","2009","2010","2011","2012","2013","2014","2015","2016","2017","2018","2019")
    axis(1,at=seq(2000,2019,1),labels=year.text)
    abline(v=seq(2000,2019,1),lty = 6, col = "cornsilk2")
    grid (NA,NULL, lty = 6, col = "cornsilk2")
    
    
    legend('topleft',legend=variableList[[4]],
           col=c("green","blue","purple","orange","dark green"),lty=c(1,1,2,2,2,3),lwd=c(2,2,2,2,2,3),cex=1,bty ="n")
    text(2012.5,4,paste0(osc[j]," oscillation: Explained variance = ",var_explained[[j]][[4]],"%"),cex=1,font = 1)
}  
```

## Cluster analysis

A heirarchical cluster anaysis was conducted on the groundwater levels in the wells. This resulted in a groups division of wells.

```{r}
# Prepare data
W <- read.csv("Data_Processing/GWLevel_imputed.csv", sep=",")[-1]
S <- read.csv("Data_Processing/SW_imputed.csv", sep=",")[-1]
P <- read.csv("Data_Processing/Precipitation_imputed.csv", sep=",")[-1]
Tm <- read.csv("Data_Processing/Temperature_imputed.csv", sep=",")[-1]
Pu <- -read.csv("Data_Processing/Pumping.csv", sep=",")[-1]

#Combining all variables
All <- data.frame(W,S,P,Tm)
All$Pu <- c(unlist(Pu),rep(NA, nrow(All)-nrow(Pu)))
head(All)

#Selecting variables for cluster analysis
onlyWells <- scale(All[,1:12])
allVar <- scale(All[1:156,])
```
```{r fig.height=5,fig.width=10}
#Using Pearson's correlation coeffiecint for oly groundwater levels
cordat<-diss(t(onlyWells), "COR")
hc<-hclust(cordat, method="ward.D")
plot(hc,main=" ",sub =" ",xlab = " ")
rect(0.7,-0.15,6.3,1.3,border =  "red")
rect(6.6,-0.15,9.1,1.3,border =  "red")
rect(9.5,-0.15,12.2,1.35,border =  "red")
mtext('Group I',side = 1,at = 3)
mtext('Group III',side = 1,at = 8)
mtext('Group II',side = 1,at = 11)
```

Now, all the variables can also be plotted to understand the relationship between the variables.

```{r fig.height=5,fig.width=10}
cordat<-diss(t(allVar), "COR")
hc<-hclust(cordat, method="ward.D")
plot(hc,main=" ",sub =" ",xlab = " ")
rect.hclust(hc, k=3,border = 'red', )
mtext('C1',side = 1,at = 8)
mtext('C2',side = 1,at = 25)
mtext('C3',side = 1,at = 32)
```

These groups obtained from cluster analysis can be compared using decadal ans seasonal oscillations from MSSA.

```{r echo=FALSE}
## Read original, decadal, and seasonal dataframes
OT <- scale(read.csv("Data_Processing/GWLevel_imputed.csv"),scale=F)[,-1]
DT <- scale(read.csv("Approach_II/Decadal_RC_of_groundwater.csv"),scale=F)[,-1]
ST <- scale(read.csv("Approach_II/Seasonal_RC_of_groundwater.csv"),scale=F)[,-1]

wellNo <- c("W60","W63","W67","W70","W73","W74","W78","W80","W81","W115","W116","W118")
# Original time series
for(i in 1:length(wellNo)){
    assign(paste0("o","_",wellNo[i],".ts"),
           ts(OT[,i],start=c(2000,1),end=c(2018,12),frequency = 12))
    
}
# Decadal time series
for(i in 1:length(wellNo)){
    assign(paste0("d","_",wellNo[i],".ts"),
           ts(DT[,i],start=c(2000,1),end=c(2018,12),frequency = 12))
    
}
# Seasonal time series
for(i in 1:length(wellNo)){
    assign(paste0("s","_",wellNo[i],".ts"),
           ts(ST[,i],start=c(2000,1),end=c(2018,12),frequency = 12))
    
}
```


```{r echo =FALSE,fig.height=5,fig.width=10}
## Plot all decadal and seasonal based on groups of groundwater wells
par(mfcol=c(3,2), mar=c(0,3.5,0,1), oma=c(4,2,2,0), mgp=c(2,.6,0), cex.lab=1.1, tcl=-.3, las=1)
#Group1
ts.plot(d_W63.ts,d_W67.ts,d_W78.ts,d_W81.ts,d_W115.ts,d_W118.ts,
        gpars = list(xlab=" ",ylab=" ",cex=1,xaxt="n",yaxt="n"),
        col=c("blue","purple","orange","green","dark green","maroon"),lwd=c(2,2,2,2),lty=c(1,1,1,1,1,1),ylim=c(-2,2.5))
abline(v=seq(2000,2019,1),lty = 6, col = "cornsilk2")
axis(2,cex.axis=1.2)
grid (NA,NULL, lty = 6, col = "cornsilk2")
legend('topleft',horiz = TRUE,legend=c("W63","W67","W78","W81","W115","W118"),
       col=c("blue","purple","orange","green","dark green","maroon"),lwd=c(2,2,2,2),lty=c(1,1,1,1,1,1,1),cex=1.2,bty="y")
text(2018,-1.8,"Group I",cex=1.5,bty="n")



#Group3
ts.plot(d_W70.ts,d_W73.ts,d_W116.ts,
        gpars = list(xlab=" ",ylab=" ",cex=1,xaxt="n",yaxt="n"),
        col=c("maroon","dark green","purple","orange","blue","green"),lwd=c(2,2,2,2),lty=c(1,1,1),ylim=c(-2,2.5))
abline(v=seq(2000,2019,1),lty = 6, col = "cornsilk2")
axis(2,cex.axis=1.2)
grid (NA,NULL, lty = 6, col = "cornsilk2")
legend('topleft',horiz = TRUE,legend=c("W70","W73","W116"),
       col=c("maroon","dark green","purple","orange","blue","green"),lwd=c(2,2,2,2),lty=c(1,1,1),cex=1.2,bty="y")
text(2018,-1.8,"Group II",cex=1.5,bty="n")

#Group2
ts.plot(d_W60.ts,d_W74.ts,d_W80.ts,
        gpars = list(xlab=" ",ylab=" ",cex=1,xaxt="n",yaxt="n"),
        col=c("maroon","green","blue","purple","blue","green"),lwd=c(2,2,2,2),lty=c(1,1,1),ylim=c(-2,2.5))
abline(v=seq(2000,2019,1),lty = 6, col = "cornsilk2")
axis(2,cex.axis=1.2)
grid (NA,NULL, lty = 6, col = "cornsilk2")
legend('topleft',horiz = TRUE,legend=c("W60","W74","W80"),
       col=c("maroon","green","blue","purple","blue","green"),lwd=c(2,2,2,2),lty=c(1,1,1),cex=1.2,bty="y")
text(2018,-1.8,"Group III",cex=1.5,bty="n")

year.text=c("2000","2001","2002","2003","2004","2005","2006","2007","2008","2009","2010","2011","2012","2013",
            "2014","2015","2016","2017","2018","2019")
axis(1,at=seq(2000,2019,1),labels=year.text,cex.axis = 1.2)
mtext("Standardized units",side = 2,outer = TRUE,las = 0,cex = 1.2)
mtext("Year",side = 1,outer = TRUE,line = 2,cex = 1.2)


## Seasonal time series
#Group1
ts.plot(s_W63.ts,s_W67.ts,s_W78.ts,s_W81.ts,s_W115.ts,s_W118.ts,
        gpars = list(xlab=" ",ylab=" ",cex=1,xaxt="n",yaxt="n"),
        col=c("blue","purple","orange","green","dark green","maroon"),lwd=c(2,2,2,2),lty=c(1,1,1,1,1,1),ylim=c(-1.5,1.5))
abline(v=seq(2000,2019,1),lty = 6, col = "cornsilk2")
axis(2,cex.axis=1.2)
grid (NA,NULL, lty = 6, col = "cornsilk2")
legend('topleft',horiz = TRUE,legend=c("W63","W67","W78","W81","W115","W118"),
       col=c("blue","purple","orange","green","dark green","maroon"),lwd=c(2,2,2,2),lty=c(1,1,1,1,1,1,1),cex=1.2,bty="y")
text(2018,-1.3,"Group I",cex=1.5,bty="n")



#Group2
ts.plot(s_W70.ts,s_W73.ts,s_W116.ts,
        gpars = list(xlab=" ",ylab=" ",cex=1,xaxt="n",yaxt="n"),
        col=c("maroon","dark green","purple","orange","blue","green"),lwd=c(2,2,2,2),lty=c(1,1,1),ylim=c(-1.5,1.5))
abline(v=seq(2000,2019,1),lty = 6, col = "cornsilk2")
axis(2,cex.axis=1.2)
grid (NA,NULL, lty = 6, col = "cornsilk2")
legend('topleft',horiz = TRUE,legend=c("W70","W73","W116"),
       col=c("maroon","dark green","purple","orange","blue","green"),lwd=c(2,2,2,2),lty=c(1,1,1),cex=1.2,bty="y")
text(2018,-1.3,"Group II",cex=1.5,bty="n")

#Group3
ts.plot(s_W60.ts,s_W74.ts,s_W80.ts,
        gpars = list(xlab=" ",ylab=" ",cex.axis=1.2,xaxt="n",yaxt="n"),
        col=c("maroon","green","blue","purple","blue","green"),lwd=c(2,2,2,2),lty=c(1,1,1),ylim=c(-1.5,1.5))
abline(v=seq(2000,2019,1),lty = 6, col = "cornsilk2")
axis(2,cex.axis=1.2)
grid (NA,NULL, lty = 6, col = "cornsilk2")
legend('topleft',horiz = TRUE,legend=c("W60","W74","W80"),
       col=c("maroon","green","blue","purple","blue","green"),lwd=c(2,2,2,2),lty=c(1,1,1),cex=1.2,bty="y")
text(2018,-1.3,"Group III",cex=1.5,bty="n")

year.text=c("2000","2001","2002","2003","2004","2005","2006","2007","2008","2009","2010","2011","2012","2013",
            "2014","2015","2016","2017","2018","2019")
axis(1,at=seq(2000,2019,1),labels=year.text,cex.axis = 1.2)
```