---
title: "Approach II"
author: "Astha Bista"
date: "2/15/2021"
output: md_document
bibliography: [packages.bib,ref.bib]
biblio-style: "apalike"
link-citations: true
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r , warning = FALSE, include=FALSE, echo=FALSE}
list.of.packages <- c('Rssa')
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
library(Rssa)

```

## Preparing the set of variables
The first task in Approach II is to set subgroups of variables such as groundwater wells, gauge stations, weather stations and pumping, grouping the stations that are close in proximity grographically. Here, we had a group of 12 sets of variables based on the location of each ground water wells. In this document, only the code for one subgroup is being shown since the codes for all the groups are similar.

```{r}
#Gather all required variable files
GW<-read.csv(file ="Data_Processing/GWLevel_imputed.csv",header = TRUE) [-1] 
SW<-read.csv(file ="Data_Processing/SW_imputed.csv",header = TRUE) [-1] 
Pr<-read.csv(file ="Data_Processing/Precipitation_imputed.csv",header = TRUE)[-1] 
Tmp<-read.csv(file ="Data_Processing/Temperature_imputed.csv",header = TRUE)[-1] 
Pmp<-read.csv(file ="Data_Processing/Pumping.csv",header = TRUE)[-1]


#The sets are named on the basis of groundwater well in the area
w60Gp <- data.frame(GW[,1],SW[,2],Pr[,3],Tmp[,3],c(unlist(Pmp),rep(NA, nrow(GW)-nrow(Pmp)))) ; #W60,G42,KY
colnames(w60Gp) <- c("W60","G42","KY_P","KY_T","Pumping") 
head(w60Gp)
```

## MSSA operation

The above is the data containing data from the subgroup containing W60 well. This data is then used for MSSA operation, for which guidance from [@golyandina2013multivariate] and [@hassani2013multivariate] were taken. Rssa package [@Rssa2015] was used for multivariate singular spectrum analysis (MSSA).

```{r}
Data<-scale(read.csv(file ="Approach_II/MSSA_Work/W60_variables_set.csv",header = TRUE)[-1])
```

Selecting a window length is somewhat of an iterative process where we select a window length and check if it gives us better separability result. Here, following suggestion in [@golyandina2013multivariate], a window length close to half of the number of observations, and a number divisible by the cycle length (12 in this case) was taken.

```{r}
L=108 #Assume window length close to N/2 (N=222 here)
#Create a ssa object
s <- ssa(Data, L = L, kind = "mssa")
```

The eigenvectors from the SVD decomposition in MSSA can be plotted to give and idea on how to group them later according to type and frequency of the vector.

```{r}
par(mfrow=c(2,2))
perc <- c("18.72%","18.53%","10.08%","7.75%")
for(i in 1:4){
        plot(s$U[,i], type="l",ylab =" ",
             xlab = "Months",main=paste0("F",i,": ",perc[i]),cex.main=1)
}
```

The separability of eigenvectors can also be assessed by plotting w correlation graph, which is defined by [@golyandina2013multivariate]. 

```{r}
# Calculate the w-correlation matrix between first 20 series
# for a guess for grouping
w <- wcor(s, groups = 1:20)
plot(w, grid = c(2,4, 5,7))
```

Using this plot, we can find which eigenvectors have high correations with each other and can be grouped together to form reconstructed components (RCs). Here, 10 groups of eigenvectors were selected for reconstruction of components. Each of these 10 reconstructed components is supposed to represent a unique characteristics that were embedded in the multivariate time series.

```{r}
r <- reconstruct(s,groups = list(c(1,2),c(3,4),c(5,6),c(7,10),c(8,9),c(11),c(12,14),
                                 c(13,15,18),c(16,17),c(19,20)))

```

Here is a plot comparing the decadal and seasonal oscillations in the variables of the subgroup of W70 well.
```{r echo=FALSE}
variableList <- list(c("W60","G42","Kearney (KY) precipitation","Kearney (KY) temperature","Pumping"),
                     c("W63","G42","Kearney (KY) precipitation","Kearney (KY) temperature","Pumping"),
                     c("W67","G45","Kearney (KY) precipitation","Kearney (KY) temperature","Pumping"),
                     c("W70","G40","Canaday steam plant (CDY) precipitation","Canaday steam plant (CDY) temperature","Pumping"), c("W73","G40","Canaday steam plant (CDY) precipitation","Canaday steam plant (CDY) temperature","Pumping"),
                     c("W74","G42","Gothenberg (GO) precipitation","Gothenberg (GO) temperature","Pumping"),
                     c("W78","G49","Grand Island (GI) precipitation","Grand Island (GI) temperature","Pumping"),
                     c("W80","G42","Gothenberg (GO) precipitation","Gothenberg (GO) temperature","Pumping"),
                     c("W81","G49","Grand Island (GI) precipitation","Grand Island (GI) temperature","Pumping"),
                     c("W115","G53","Grand Island (GI) precipitation","Grand Island (GI) temperature","Pumping"),
                     c("W116","G53","Grand Island (GI) precipitation","Grand Island (GI) temperature","Pumping"),
                     c("W118","G65","David City (DC) precipitation","David City (DC) temperature","Pumping"))

# List of seasonal and decadal variations in MSSA operations for each respective subgroup
var_explained <- list(c(36.08,33.92,35.65,37.84,38.53,36.01,38.47,37.67,38.66,37.72,37.93,38.06),
                      c(18.19,27.76,23.62,24.8,24.34,19.57,23.8,17.64,23.69,23.35,21.71,15.8))
osc <- c("Seasonal","Decadal")

list <- c("W","G","P","T","Pu")
```

```{r fig.height=5,fig.width=15}

par(mfrow = c(1, 2))

## Loop around seasonal and decadal oscillations
for (j in 1:2){
    # Assign filepath for each subgroup
    assign(paste0("filepath",j),Sys.glob(path=paste0("Approach_II/MSSA_Work/W70_RC",
                                                     sprintf("%02d",as.numeric(j)),".csv")))
    # Read RC files for decadal and seasonal in each subgroup
    assign(paste0("F",j),read.csv(file=get(paste0("filepath",j)),header = TRUE))
    
    ## Loop around each variable 
    for (k in 1:length(list)){
        # Create time series object for each variable
        assign(paste0(list[k],".ts"),ts(get(paste0("F",j))[,k+1],start=c(2000,1),end=c(2018,12),
                                        frequency = 12))
    }
    
    ## Plot time series of all variables for decadal and seasonal separately
    ts.plot(W.ts,G.ts,P.ts,T.ts,Pu.ts,
            gpars = list(xlab="Year",ylab="Standardized units",xaxt="n",ylim=c(-2,4),
                         col=c("green","blue","purple","orange","dark green"),lty=c(1,1,2,2,2,3),lwd=c(2,2,2,2,2,3)))
    
    year.text=c("2000","2001","2002","2003","2004","2005","2006","2007",
                "2008","2009","2010","2011","2012","2013","2014","2015","2016","2017","2018","2019")
    axis(1,at=seq(2000,2019,1),labels=year.text)
    abline(v=seq(2000,2019,1),lty = 6, col = "cornsilk2")
    grid (NA,NULL, lty = 6, col = "cornsilk2")
    
    
    legend('topleft',legend=variableList[[4]],
           col=c("green","blue","purple","orange","dark green"),lty=c(1,1,2,2,2,3),lwd=c(2,2,2,2,2,3),cex=1,bty ="n")
    text(2012.5,4,paste0(osc[j]," oscillation: Explained variance = ",var_explained[[j]][[4]],"%"),cex=1,font = 1)
    }  
```


## References
