---
title: "ProcessingData"
author: "Astha Bista"
date: "1/18/2021"
output: md_document
bibliography: [packages.bib, ref.bib]
biblio-style: "apalike"
link-citations: true
---
```{r , echo=FALSE, warning=FALSE}
setwd("C:/Users/Aastha/Desktop/GWProject")
list.of.packages <- c("ggplot2", "forecast","reshape2","dplyr","forecast")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
```

# Groundwater levels
Groundwater levels were obtained from [United States Geological Survey (USGS) National Groundwater Monitoring Network](https://cida.usgs.gov/ngwmn/) for Platte River Basin in Nebraska from 2000 to 2018. 12 groundwater wells were selected based on data available, and screen depth. 
Read the data containing groundwater levels with missing values:

```{r rawdata, echo=FALSE}
gwlevel <- read.csv(file ="Data_Processing/GWLevel_with_missinggaps.csv",header = TRUE)
head(gwlevel)
```

Count the missing values
```{r na, echo=FALSE}
na_count <-sapply(gwlevel, function(y) sum(length(which(is.na(y)))))
na_count
```


## Imputation of missing values and visualization of data

The missing values in the data are imputed using na_interp command in forecast package[@R-forecast]. In this command, for seasonal series, a robust STL decomposition is first computed. Then a linear interpolation is applied to the seasonally adjusted data, and the seasonal component is added back. @moritz2015comparison compared interpolation methods and found that this method gave the least error.

```{r ,echo = FALSE, warning=FALSE}
library(forecast)
gw_imputed<-c()
for(i in 2:ncol(gwlevel)){
        imp <- na.interp(gwlevel[,i])
        gw_imputed<-cbind(gw_imputed,imp)
}
       
colnames(gw_imputed)<-c("W60","W63","W67","W70","W73","W74","W78","W80","W81","W115","W116","W118")
head(gw_imputed)
```
Here is a plot of the points interpolated in one of the time series.
```{r ,echo = FALSE,fig.height=3,fig.width=6, warning=FALSE}
dat <- gwlevel[,11]*0.3048
dates = seq(from = as.Date("2000-01-01"), to = as.Date("2018-12-1"), by = 'month')
imp <- na.interp(dat)
x <- 2000 + (which(is.na(dat))-1)/12 
y <- imp[which(is.na(dat))]

dat.ts<-ts(dat,start=c(2000,1),end=c(2018,12),frequency = 12)
ts.plot(dat.ts,gpars = list(xlab="Year",ylab="Water level (m)"),main = "Imputed values",type="o",lwd=1,col="blue")
points(x,y,col="red",pch=1,cex=1) 
mtext(colnames(gwlevel)[11])
grid()
```
```{r,echo = FALSE, warning=FALSE}
gwlevel_m <- sapply(as.data.frame(gw_imputed), function(y) y*0.3048)
```

```{r fig.height=10,fig.width=10,echo = FALSE, warning=FALSE}
dates = seq(from = as.Date("2000-01-01"), to = as.Date("2018-12-1"), by = 'month')
gwlevel_m2 <- data.frame(as.data.frame(gwlevel_m),dates)
colnames(gwlevel_m2)[13]<-"Date"
library(reshape2)
library(ggplot2)
library(dplyr)
gwlevel_m3 <- melt(gwlevel_m2, id.vars = "Date", 
                   variable.name = "Wells", value.name="WL")
gwlevel_m4 <- gwlevel_m3%>%
        group_by(Date,Wells)

ggplot(data = gwlevel_m4,aes(x = Date, y = WL)) + 
        geom_line() + facet_wrap(~ Wells, nrow = 6, scales = "free") +
        xlab("Year") + ylab("Water level (m)") + theme_bw()
```


# Stream gauge height data
Stream gauge heights were obtained from [USGS National Water Information System (NWIS)](https://waterdata.usgs.gov/nwis) for Platte River Basin in Nebraska from 2000 to 2018. There were 12 gauge stations in the study area that contained data in the aforementioned period with some missing values. 
The stream gauge height data are given such as:

```{r echo=FALSE}
swgaugeht <- read.csv(file ="Data_Processing/SW_with_missinggaps.csv",header = TRUE)
head(swgaugeht)
```
Now, the missing values in the time series are give in percentage of the number of observations in each time series.
```{r echo=FALSE}
na_count <-sapply(swgaugeht[,-1], function(y) sum(length(which(is.na(y)))))
(na_count/nrow(swgaugeht))*100
```
## Imputation of missing values and visualization of data

Out of the 12 gauge stations, seven stations were selected based on the percentage of missing values.Interpolation was carried out using na_interp to impute missing values.

```{r ,echo = FALSE, warning=FALSE}
library(imputeTS)
sw_imputed<-c()
for(i in c(2,3,6,10,11,12,13)){
        imp <- na.interp(swgaugeht[,i])
        sw_imputed<-cbind(sw_imputed,imp)
}

colnames(sw_imputed)<-c("G40","G42","G45","G49","G51","G53","G65")

head(sw_imputed)
```
Here is a plot of the points interpolated in one of the time series.
```{r ,echo = FALSE,fig.height=3,fig.width=6, warning=FALSE}
dat <- swgaugeht[,2]*0.3048
dates = seq(from = as.Date("2000-01-01"), to = as.Date("2018-12-1"), by = 'month')
imp <- na.interp(dat)
x <- 2000 + (which(is.na(dat))-1)/12 
y <- imp[which(is.na(dat))]

dat.ts<-ts(dat,start=c(2000,1),end=c(2018,12),frequency = 12)
ts.plot(dat.ts,gpars = list(xlab="Year",ylab="Water level (m)"),main = "Imputed values",
        type="o",lwd=1,col="blue")
points(x,y,col="red",pch=1,cex=1) 
mtext(colnames(swgaugeht)[2])
grid()

```
```{r,echo = FALSE}
swgaugeht_m <- sapply(as.data.frame(sw_imputed), function(y) y*0.3048)
```

Here are plots of stream gauge heights after imputations.

```{r fig.height=10,fig.width=10,echo = FALSE, warning=FALSE}
dates = seq(from = as.Date("2000-01-01"), to = as.Date("2018-12-1"), by = 'month')
swgaugeht_m2 <- data.frame(as.data.frame(swgaugeht_m),dates)
colnames(swgaugeht_m2)[8]<-"Date"
library(reshape2)
library(ggplot2)
library(dplyr)
swgaugeht_m3 <- melt(swgaugeht_m2, id.vars = "Date", 
                   variable.name = "GaugeSt", value.name="GgHt")
swgaugeht_m4 <- swgaugeht_m3%>%
        group_by(Date,GaugeSt)
swgaugeht_m4
ggplot(data = swgaugeht_m4,aes(x = Date, y = GgHt)) + 
        geom_line() + facet_wrap(~ GaugeSt, nrow = 6, scales = "free") +
        xlab("Year") + ylab("Gauge height (m)") + theme_bw()
```



# Precipitation data
Stream gauge heights were obtained from [USGS National Water Information System (NWIS)](https://waterdata.usgs.gov/nwis) for Platte River Basin in Nebraska from 2000 to 2018. There were 12 gauge stations in the study area that contained data in the aforementioned period with some missing values. 
The stream gauge height data are given such as:

```{r echo=FALSE}
prec <- read.csv(file ="Data_Processing/Precipitation_with_missinggaps.csv",header = TRUE)
head(prec)
```
Now, the missing values in the time series are give in percentage of the number of observations in each time series.
```{r echo=FALSE}
colnames(prec)<-c("RV","CDY","KY","HY","GI","GS","GO","NP","DC")
na_count <-sapply(prec, function(y) sum(length(which(is.na(y)))))
(na_count/nrow(prec))*100
```
## Imputation of missing values and visualization of data

Out of the 12 gauge stations, seven stations were selected based on the percentage of missing values. Interpolation was carried out using na_interp to impute missing values.

```{r ,echo = FALSE, warning=FALSE}
library(imputeTS)
prec_imputed<-c()
for(i in 1:ncol(prec)){
        imp <- na.interp(prec[,i])
        prec_imputed<-cbind(prec_imputed,imp)
}

colnames(prec_imputed)<-c("RV","CDY","KY","HY","GI","GS","GO","NP","DC")

head(prec_imputed)
```
Here is a plot of the points interpolated in one of the time series (units changed from inches to mm).
```{r ,echo = FALSE,fig.height=3,fig.width=6, warning=FALSE}
dat <- prec[,1]*25.4
dates = seq(from = as.Date("2000-01-01"), to = as.Date("2018-12-1"), by = 'month')
imp <- na.interp(dat)
x <- 2000 + (which(is.na(dat))-1)/12 
y <- imp[which(is.na(dat))]

dat.ts<-ts(dat,start=c(2000,1),end=c(2018,12),frequency = 12)
ts.plot(dat.ts,gpars = list(xlab="Year",ylab="Precipitation (mm)"),main = "Imputed values",
        type="o",lwd=1,col="blue")
points(x,y,col="red",pch=1,cex=1) 
mtext(colnames(prec)[7])
grid()

```
```{r,echo = FALSE}
prec_m <- sapply(as.data.frame(prec_imputed), function(y) y*25.4)
```

Here are plots of stream gauge heights after imputations.

```{r fig.height=10,fig.width=10,echo = FALSE, warning=FALSE}
prec_m <- sapply(as.data.frame(prec_imputed), function(y) y*25.4)

dates = seq(from = as.Date("2000-01-01"), to = as.Date("2018-12-1"), by = 'month')
prec_m2 <- data.frame(as.data.frame(prec_m),dates)
colnames(prec_m2)[10]<-"Date"
library(reshape2)
library(ggplot2)
library(dplyr)
prec_m3 <- melt(prec_m2, id.vars = "Date", 
                     variable.name = "Prec", value.name="Amount")
prec_m4 <- prec_m3%>%
        group_by(Date,Prec)

ggplot(data = prec_m4,aes(x = Date, y = Amount)) + 
        geom_line() + facet_wrap(~ Prec, nrow = 6, scales = "free") +
        xlab("Year") + ylab("Precipitation (mm)") + theme_bw()
```
# Pumping data
The pumping data was obtained from the [CENEB Groundwater model report](https://dnr.nebraska.gov/CENEB-Groundwater-Model) prepared by [Nebraska Department of Natural Resources](https://dnr.nebraska.gov/). The time period is from Jan 2000 to Dec 2012.


```{r echo=FALSE}
pmp <- read.csv(file ="Data_Processing/Pumping.csv",header = TRUE)
head(pmp)
```

The pumping plot is shown below. Please note that the y-axis has been reversed for purpose of comparison with water levels from other variables.
```{r echo=FALSE, warning=FALSE}
pmp_m <- pmp[,2]* 0.0283168
dates = seq(from = as.Date("2000-01-01"), to = as.Date("2012-12-01"), by = 'month')
pmp_m2 <- data.frame(as.data.frame(pmp_m),dates)
colnames(pmp_m2)<-c("Pumping","Date")
library(ggplot2)
```
```{r fig.height=3,fig.width=7,echo = FALSE}
ggplot(data = pmp_m2,aes(x = Date, y = Pumping)) + 
        geom_line() + xlab("Year") + ylab("Pumping (cubic metres)") + scale_y_reverse()+
        scale_x_date(date_breaks = "12 months",date_labels = "%Y") + theme_bw()

```

# References